---
title: "How has COVID-19 affected the US substance abuse crisis?"
author: "Chris Hanson"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  fig.width = 10
  )
```

```{r libraries, include = FALSE}
library(httr)
library(xml2)
library(data.table)
library(ggplot2)
library(tidyverse)
library(lubridate)
library(rvest)
library(stringr)
library(ggsci)
library(plotly)
library(zoo)
```

```{r state_names}
abbrev2 <- read_html(x = "https://simple.wikipedia.org/wiki/List_of_U.S._states_by_traditional_abbreviation")
abtable2 <- xml_find_all(abbrev2, xpath = "/html/body/div[3]/div[3]/div[5]/div[1]/table[1]/tbody")
abtable2 <- html_table(abtable2)
abtable2 <- rbindlist(abtable2)
abtable2 <- abtable2[, .(State, Otherabbreviations)]
setnames(abtable2, "State", "state_name")
setnames(abtable2, "Otherabbreviations", "state_abbreviation")
abtable2$state_name = toupper(abtable2$state_name)
abtable2$state_abbreviation = toupper(abtable2$state_abbreviation)
abtable2$state_abbreviation <- str_replace_all(abtable2$state_abbreviation, "[[:punct:]]", "")
abtable2[11,1] <- "GEORGIA"
abtable2 <- abtable2 %>% add_row(state_name = "UNITED STATES", state_abbreviation = "US")
abtable2 <- abtable2[order(state_name)]
```

```{r census_api}
# https://www.census.gov/data/developers/guidance/api-user-guide.Example_API_Queries.html
# Retrieving dimensions
if (!exists("censusquery")){
censusquery <- GET(
  url = "https://api.census.gov/data/2019/pep/population?get=NAME,POP&for=state:*"
)
}

# Checking if the website is active
#censusquery$status_code

# Processing data
census_content <- content(censusquery)
census <- rbindlist(census_content)
census <- census[-1,]
setnames(census, "V2", "population")
setnames(census, "V1", "state_name")
setnames(census, "V3", "state_number")
census$population <- as.numeric(census$population)
census$state_number <- as.numeric(census$state_number)
census <- census %>% add_row(state_name = "United States", population = sum(census$population), state_number = 57)
census <- census[order(state_name)]

#Combining census and abtable data.tables
census$state_name = toupper(census$state_name)
census <- merge(census, abtable2)
#census$population <- as.numeric(census$population)
#census$population <- as.numeric(census$population)
```

```{r regions}
divisions <- fread('data/us_divisions.csv')
divisions$State = toupper(divisions$State)

# merge with census

census <- merge(census, divisions, by.x = "state_name", by.y = "State", all.x = TRUE)
census$'State Code' <- NULL
census$Division <- NULL
setnames(census, "Region", "region")
```

```{r covid_api}
# Retrieving dimensions
if(!exists("covidquery")){
covidquery <- GET(
  url = "https://data.cdc.gov/resource/9mfq-cb36.json?$limit=50000"
)
}

# Checking if the website is active
#covidquery$status_code

# Processing data
covid_content <- content(covidquery)

covid <- rbindlist(covid_content[1:length(covid_content)], fill=TRUE)
covid$date <- as.Date(covid$submission_date)
covid$tot_death <- as.numeric(covid$tot_death)
covid$tot_cases <- as.numeric(covid$tot_cases)
covid$new_case <- as.numeric(covid$new_case)
covid$new_death <- as.numeric(covid$new_death)

# Removing extraneous columns from 'covid'
covid <- covid[, .(state, tot_cases, new_case, tot_death, new_death, date)]
covid <- covid[order(state, date)]
```

```{r merge_census_covid}
#Combining census and covid data.tables
covid <- merge(covid, census, by.x = "state", by.y = "state_abbreviation")
covid$population <- as.numeric(covid$population)
```

```{r covid_population_norming}
# Creating death_perc, total death / total population
covid$death_perc <- covid$tot_death * 100 / covid$population
covid$case_perc <- covid$tot_cases * 100 / covid$population
covid$newdeath_perc <- covid$new_death * 100 / covid$population
covid$newcase_perc <- covid$new_case * 100 / covid$population

# Summing up COVID data from all states, for total US values
covid_usa <- aggregate(cbind(tot_cases, new_case, tot_death, new_death, death_perc, case_perc, newdeath_perc, newcase_perc, population) ~ date, data = covid, FUN = sum, na.rm = TRUE)
```

```{r covid_graph_by_state}
# Plotting covid deaths by population percent
ccpp <- ggplot(covid, aes(x = date, y = case_perc, color = state_name)) +
  geom_line() +
  labs(title = "COVID-19 cases by % of state population", subtitle = "Source: data.cdc.gov API") +
  xlab("Date") + ylab("COVID-19 cases by % of state population") +
  theme(legend.title = element_text(size = 5), legend.text = element_text(size = 5))

cdpp <- ggplot(covid, aes(x = date, y = death_perc, color = state_name)) +
  geom_line() +
  labs(title = "COVID-19 deaths by % of state population", subtitle = "Source: data.cdc.gov API")+
  xlab("Date") + ylab("COVID-19 deaths by % of state population")
```

```{r covid_max}
covid_max <- covid[ , .SD[which.max(date)], by = state]
covid_max <- covid_max[order(-death_perc)]
```

```{r drugs_api}
# Retrieving dimensions
# https://data.cdc.gov/NCHS/VSRR-Provisional-Drug-Overdose-Death-Counts/xkb8-kh2a

if(!exists("drugsquery")){
drugsquery <- GET(url = "https://data.cdc.gov/resource/xkb8-kh2a.json?$limit=50000")
}
# Checking if the website is active
#drugsquery$status_code

# Processing data
drugs_content <- content(drugsquery)

# Turning the list into a data.table
drugs <- rbindlist(drugs_content[1:length(drugs_content)], fill=TRUE)
drugs$state_name = toupper(drugs$state_name)
```

```{r drugs-processing2}
# Cleaning up the drugs data.table
drugs$date <- as.Date(paste0(drugs$year, drugs$month, "01"), format="%Y%B%d")
setnames(drugs, "data_value", "deaths")
drugs$deaths <- as.numeric(drugs$deaths)

# Combining drugs and census data
drugs <- merge(drugs, census, by = "state_name", all = TRUE)

# Removing extraneous columns from 'drugs'
drugs <- drugs[, .(state_name, state, indicator, deaths, date, year, population, state_number, region)]

# Cleaning up drugs$indicator column
drugs$indicator <- gsub("\\([^()]*\\)", "", drugs$indicator)
drugs$indicator <- str_trim(drugs$indicator, side = c("both"))
drugs <- drugs[indicator != "Percent with drugs specified"]

# Making a deaths per population column
drugs$odpop_perc <- drugs$deaths * 100 / drugs$population

# Dropping rows where deaths == NA
drugs <- drugs %>% drop_na(deaths)
drugs <- drugs[state != "DC"]
drugs <- drugs[order(state_number, date)]
```

Drug overdose deaths in the United States have been rising steadily since the turn of the century, and a significant increase in this trend has been observed since the mid-2010's. In March 2020, the COVID-19 pandemic led to an upheaval in nearly every aspect of daily life, resulting in drastic changes to the way we work, socialize, and interact with society at large. To slow the spread of the virus, most public places of congregation were shut down, leading to widespread loss of jobs and a crash of the economy.

Substance abuse is associated with unemployment or underemployment, lack of career opportunities, social isolation, mental health issues, and homelessness. As COVID-19 has contributed to each of these factors, an investigation into the pandemic’s effect on the substance abuse crisis is warranted.

<br>

## COVID-19 in the US

The COVID-19 pandemic was experienced differently in each state:

```{r}
ggplotly(cdpp)
```

The state with the highest number deaths per population, `r covid_max[1,state_name]`, has experienced per population mortality rates `r format(round(covid_max[1,death_perc] * 100 / covid_max[nrow(covid_max),death_perc]), 2)`% higher than that of the lowest number of deaths per population, `r covid_max[nrow(covid_max),state_name]`.

<br>

## The drug mortality crisis in the US

Shown below is overdoses in the US from 2015-2021, broken down by drug type and sub-type.

```{r drugs_usa_plot}
# Filtering DF to be just US
drugs_usa <- drugs %>% filter(state == "US")
drugs_usa <- drugs_usa[order(date, indicator)]

drugs_usa <- drugs_usa[!indicator %in% c("Number of Deaths", "Natural, semi-synthetic, & synthetic opioids, incl. methadone", "Natural & semi-synthetic opioids, incl. methadone", "Methadone")]
drugs_usa$region <- NULL

# Plot all indicators
gdu <- ggplot(drugs_usa, aes(x = date, y = deaths, color = indicator)) +
  geom_line(size = 1.5) +
  geom_vline(xintercept=as.Date("2020-02-01"), linetype="dashed", size = 1) +
  annotate(x = as.Date("2019-11-01"), y = 70000, label = "Start of COVID-19", geom="text", angle = 90, size = 6) +
  theme(legend.title = element_text(size = 8), legend.text = element_text(size = 6)) +
  labs(x = "Date", y = "12 month rolling sum of deaths from OD", title = "Total US deaths by OD from significant drug categories, 2015-2021") +
  scale_x_date(date_breaks = "year",  date_labels = "%Y") +
  theme(legend.title = element_text(size = 8), legend.text = element_text(size = 7)) 

gdu
```

There does appear to be a clear acceleration in the rate of drug overdose deaths following the onset of the COVID-19 pandemic. 

This plot makes clear that the drug most heavily responsible for all drug overdose deaths, as well as the most recent spike in deaths, is opioids - namely, synthetic opioids. According to numerous government agencies, the synthetic opioid behind this rise in fatalities is fentanyl, which is often used to lace other drugs. As little as two milligrams of fentanyl can be lethal.

```{r drugs_usa_2_3}
indicators = unique(drugs$indicator)
indicators2 <- c(indicators[3], indicators[6], indicators[7], indicators[8], indicators[9], indicators[11])
indicators3 <- c(indicators[2], indicators[7], indicators[3], indicators[8])

drugs_usa2 <- drugs_usa[indicator %in% indicators2]
drugs_usa3 <- drugs_usa[indicator %in% indicators3]
```

```{r drugs_usa3_plot}
gdu3 <- ggplot(drugs_usa3, aes(x = date, y = deaths, color = indicator)) +
  geom_line(size = 1.4) +
  geom_vline(xintercept=as.Date("2020-02-01"), linetype="dashed", size = 1) +
  labs(x = "Date", y = "12 month rolling sum of OD deaths by drug category", title = "Opioid contribution to total overdose deaths") +
  theme(legend.title = element_text(size = 8), legend.text = element_text(size = 7)) 
```

```{r max_min_date_drugs}
# the most recent values in the drugs dataset
maxod <- drugs_usa[date == "2021-04-01" & indicator == "Number of Drug Overdose Deaths"]
maxsynth <- drugs_usa[date == "2021-04-01" & indicator == "Synthetic opioids, excl. methadone"]

minod <- drugs_usa[date == "2015-01-01" & indicator == "Number of Drug Overdose Deaths"]
minsynth <- drugs_usa[date == "2015-01-01" & indicator == "Synthetic opioids, excl. methadone"]
```

In January 2015 synthetic opioids accounted for `r format(round(minsynth$deaths * 100 / minod$deaths, 0))`% of all overdose deaths. As of March 01, 2021 synthetic opioids were responsible for `r format(round(maxsynth$deaths * 100 / maxod$deaths, 0))`% of all drug overdose deaths in the US. During this time, the total number of annual deaths due to synthetic opioids grew by `r format(round(maxsynth$deaths * 100/ minsynth$deaths, 0))`%.

Each state is experiencing the drug epidemic in its own way:

```{r od_deaths_states_plots}
# Making a data.table of just "Number of Drug Overdose Deaths"
od_deaths <- drugs[indicator == "Number of Drug Overdose Deaths"]
od_deaths <- od_deaths[state_name != "UNITED STATES"]
od_deaths <- od_deaths[order(state_name, date)]

# Plotting od deaths scaled by population
allodpop <- ggplot(od_deaths, aes(x = date, y = odpop_perc, color = state_name)) +
  geom_line() + #show.legend = FALSE
  labs(title = "12 month rolling sum of OD deaths, by % of state pop", x = "Date", y = "12 month rolling sum of OD deaths, by % of state pop.", subtitle = "Source: data.cdc.gov API")

```

```{r plot_odstates}
ggplotly(allodpop)
```

```{r percent-increase-drugs-by-state}
od_min <- od_deaths[ , .SD[which.min(date)], by = state_name] #2015-01-01
od_realmax <- od_deaths[ , .SD[which.max(date)], by = state_name]
od_max <- od_deaths[date == "2021-04-01"] #Let's just keep this here
od_0420 <- od_deaths[date == "2020-04-01"]

# Percent increase from March 2020 to March 2021
od_max$odincrease_apr20apr21 <- (od_max$deaths - od_0420$deaths) * 100 / od_0420$deaths
```

```{r covid-vs-drugs plots}
# Covid data from 03.21 with OD % increase data from 04.20-04.21
covid_0421 <- covid[date == "2021-04-01"]
covid_od <- merge.data.frame(covid_0421, od_max)

od_cov1 <- ggplot(covid_od, aes(x = death_perc, y = odincrease_apr20apr21, label = state_name)) +
  geom_point(aes(color = region)) +
  geom_smooth() +
  labs(title = "Deaths by COVID vs deaths by overdose in US states", x = "COVID-19 deaths, by % of state pop", y = "% increase in deaths by OD, 04/20-04/21", subtitle = "Source: data.cdc.gov API")

od_cov2 <- ggplot(covid_od, aes(x = case_perc, y = odpop_perc, label = state_name)) +
  geom_point(aes(color = region)) +
  geom_smooth() +
  labs(title = "Cases of COVID vs deaths by overdose in US states", x = "% of state pop infected by COVID-19", y = "% of total state population killed by OD, April 2020 - April 2021", subtitle = "Source: data.cdc.gov API")
```

The plot below probes the question of whether statewide COVID-19 *deaths* per capita is correlated with the *percentage increase (over subsequent 12 month periods) in deaths due to overdose* per capita.

```{r plot_covidvsod}
ggplotly(od_cov1)
```

This indicates that there is either no relation between the two factors, or there is even a negative correlation. 

In an attempt to make sense of this apparent negative correlation, the question is asked whether states which experienced the lowest COVID-19 mortalities endured the most severe lockdowns - resulting in lower virus associated deaths but higher social isolation and a more damaged economy, and thus possibly more deaths due to overdose.

To investigate this possibility, another dataset will be introduced into this analysis: monthly unemployment data in the 50 US states.

<br>

## COVID-19 and unemployment in the US

To examine the possibility that the unemployment experienced as a result of the COVID-19 lockdowns led to increased drug usage and deaths by overdose, below is a visualization of the relationship between these factors: The *relative increase in the percent of the total state population that died due to drug overdose in the 12 months ending in April 2020 vs April 2021*, compared to the *increase in the percent of the state population experiencing unemployment from February 2020, before the pandemic, to October 2020,* when unemployment began to recede in most states.

```{r unemployment_fread}
# https://www.bls.gov/lau/data.htm

unemployment <- fread("data/stateunemployment.csv")
unemployment <- head(unemployment, -3)
unemployment <- melt(unemployment, id.vars="STATE", variable.name = "date")

UD <- unemployment$date
UDF <- as.Date(paste0("01-", UD), format = "%d-%b-%y")
unemployment$date <- UDF
unemployment <- unemployment[order(STATE, date)]
```

```{r unemployment_investigation}
unemployment_plot <- unemployment %>% filter(date >= "2018-01-01")
unem <- ggplot(unemployment_plot, aes(x = date, y = value, color = STATE)) +
  geom_line() +
  labs(title = "Unemployment %, per state", x = "Date", y = "Unemployment rate (seasonally adjusted)", subtitle = "Source: US Bureau of Labor Statistics")
```

```{r unemployment_cleanup}
umax <- unemployment %>% filter(date == "2020-10-01") #vary this, 10 is good
umin <- unemployment %>% filter(date == "2020-02-01") #invariable, 02 or 03
unemployment_change <- select(umin, c("STATE"))
unemployment_change$unemp_change <- umax$value - umin$value
unemployment_change$umax_value <- umax$value

covid_od_unemp <- merge(covid_od, unemployment_change, by.x = "state_name", by.y = "STATE")
covid_od_unemp <- covid_od_unemp %>% filter(!state_name %in% c("HAWAII"))
```

```{r plots_unempod}
u5 <- ggplot(covid_od_unemp, aes(x = umax_value, y = odpop_perc, label = state_name, color = death_perc)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Unemployment vs deaths by OD", x = "Unemployment %, October 2020", y = "OD deaths by % of state population, April 2020 - April 2021", color = "Deaths from\nCOVID-19 by\n% state \npopulation")

u4 <- ggplot(covid_od_unemp, aes(x = unemp_change, y = odpop_perc, label = state_name)) +
  geom_point(aes(color = region)) +
  geom_smooth() +
  labs(title = "Increase in unemployment vs deaths by OD", x = "Relative increase in unemployment, April 2020 - October 2020", y = "OD deaths by % of state population, February 2020 - April 2021")

u3 <- ggplot(covid_od_unemp, aes(x = unemp_change, y = odincrease_apr20apr21, label = state_name)) +
  geom_point(aes(color = region)) +
  geom_smooth() +
  labs(title = "Unemployment rise vs 12 month OD increase", x = "Relative increase in unemployment, February 2020 - October 2020", y = "% increase in deaths by OD from 04/19-04/20 to 04/20-04/21")
```

```{r plot_umpod3}
ggplotly(u3, tooltip = c("label"))
```

This plot shows normalized unemployment values vs the percentage increase in drug overdoses per capita for the 12 month period coinciding with COVID-19, compared to the 12 month period immediately prior. A positive correlation is suggested, but is not definitive.

<br>

## COVID-19 and economic shutdown in the US

Another way to measure the economic health of a community is to consider the change in state GDP in the period following the COVID lockdown.

Below is a visualization of the relationship between the normalized percentage drop in GDP per state during the economic shutdown vs the normalized percentage increase in deaths due to overdose during COVID-19:

```{r econ_api}
# Retrieving dimensions
# https://apps.bea.gov/api/_pdf/bea_web_service_api_user_guide.pdf, pg 11

if(!exists("econquery")){
econquery <- GET(url = "https://apps.bea.gov/api/data/?&UserID=0BACEFA6-DE0F-459B-B599-D4B53A5B7528&method=GetData&datasetname=Regional&TableName=SQGDP9&LineCode=1&GeoFips=STATE&Frequency=Q&Year=2018,2019,2020,2021&ResultFormat=JSON"
                 )} #Real GDP by state

# Checking if the website is active
#econquery$status_code

# Processing data
econ_content <- content(econquery)

# Turning the list into a data.table
econ <- rbindlist(econ_content$BEAAPI$Results$Data[1:length(econ_content$BEAAPI$Results$Data)], fill=TRUE)

econ$TimePeriod <- as.Date(as.yearqtr(econ$TimePeriod))
econ$DataValue <- as.numeric(gsub(",","",econ$DataValue))
econ$GeoName <- toupper(econ$GeoName)

econ <- econ %>% filter(GeoName %in% abtable2$state_name)
econ <- econ %>% filter(GeoName != "UNITED STATES")
```

```{r econ_cleanup}
# Minimum options for comparing relative GDP change
#econ_min <- econ[ , .SD[which.min(TimePeriod)], by = GeoFips]
econ_min <- econ[TimePeriod == "2019-10-01"]
setnames(econ_min, "DataValue", "relative_value")

econ_relative <- merge(econ, econ_min, by = "GeoName")
econ_relative$relative <- econ_relative$DataValue / econ_relative$`relative_value`
#econ_relative <- filter(econ_relative, TimePeriod.x >= "2019-07-01")
econ_relative <- econ_relative[order(GeoName, TimePeriod.x)]
```

```{r econ_plot}
ggecon <- ggplot(econ, aes(x = TimePeriod, y = DataValue, color = GeoName)) +
  geom_point() +
  labs(title = "Quarterly state GDP", x = "Date", y = "State GDP (millions of chained 2012 dollars)", subtitle = "Source: US Bureau of Economic Analysis API")

ggecon_rel <- ggplot(econ_relative, aes(x = TimePeriod.x, y = relative, color = GeoName)) +
  geom_point() +
  labs(title = "Quarterly state GDP, relative to 4th quarter 2019", x = "Date", y = "State GDP (relative to 4th qtr 2019)", subtitle = "Source: US Bureau of Economic Analysis API")
```

```{r econ_plot2}
q1 <- econ %>% filter(TimePeriod == "2020-04-01") #low value, post-covid
q2 <- econ %>% filter(TimePeriod == "2019-10-01") #high value, pre-covid
q3 <- (1 - (q1$DataValue / q2$DataValue)) * 100 #total percentage drop in GDP

econ_drop <- econ %>% filter(TimePeriod == "2021-04-01")
econ_drop$econdrop <- q3
econ_drop <- select(econ_drop, c("GeoName", "econdrop"))
covid_od_econ <- merge(covid_od, econ_drop, by.x = "state_name", by.y = "GeoName")
covid_od_econ <- covid_od_econ %>% filter(!state_name %in% c("HAWAII", "ALASKA"))

w1 <- ggplot(covid_od_econ, aes(x = econdrop, y = odincrease_apr20apr21, label = state_name, color = case_perc)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Economic recession vs 12 month OD increase", x = "% drop in state GDP, QTR4 2019 - QTR2 2020", y = "% increase in deaths by OD, 04/19-04/20 vs 04/20-04/21", color = "Cases of\nCOVID-19 by\n% state population")

ggplotly(w1, tooltip = c("label"))
```

<br>

## Conclusion and Summary

This project has been an attempt to sift through the available datasets and detect incontrovertible relationships between the indicators in question – COVID-19 cases and associated deaths, and deaths due to drug overdose. As these crises both affect and are affected by the common circumstances of daily life - employment, health, mental wellbeing - the question of whether they are directly influencing each other is a natural one. Ultimately, the analyses performed did suggest a correlation between these variables and were intriguing enough to warrant further investigation, but were not indisputable beyond all doubt. 


